Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-04-28 Manuscript/PLoSONESubmission/TEM_PLoS_draft.tex

% Please do not include colors or graphics in the text.
%
% Please do not create a heading level below \subsection. For 3rd level headings, use \paragraph{}.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout or alignment
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment.
%
\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}
% Use Unicode characters when possible
\usepackage[utf8]{inputenc}
% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}
% fixltx2e package for \textsubscript
\usepackage{fixltx2e}
% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}
% line numbers
\usepackage[right]{lineno}
% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
\usepackage{rotating}
\usepackage{setspace} 
\doublespacing

@@

We inferred the morphological matrix using the R package ape v. 3.0-11 \cite{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (see Appendix S1 and Fig. S1). We then ran an independent discrete character simulation for each character using the "true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \cite{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \cite{wagner2000,davalosintegrating2014,wrightbayesian2014}.

@@

Each Bayesian tree was estimated using two runs of four chains each for a maximum of 5$\times$$10^7$ generations. We used the following priors for each tree: (i) the "true" tree’s topology as a starting tree (with a starting value for each branch length of one), (ii) an exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions, and (iii) a transition/transversion ratio prior of two sampled from a strong beta distribution ($\beta$(80,40)). We used these priors to speed up the Bayesian estimation process. These priors biased the way the Bayesian process calculated branch lengths by giving non-random starting points and boundaries for parameter estimation however, here we are focusing on the effect of missing data on tree topology and not branch lengths. Even using these priors, it took \texttildelow 140 CPU years to build 50 sets of 125 Bayesian trees (2.30GHz clock speed nodes). The detailed MrBayes parameters are available in the Appendix S1.

@@

\paragraph*{Robinson-Foulds distance}
The Robinson-Foulds distance \cite{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees \cite{RF1981} (see Appendix S2 for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric by dividing the Robinson-Foulds distance by the mean Robinson-Foulds distance from 1000 random tree comparisons \cite{Bogdanowicz2012}. This allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see Appendix S2 for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the distance is equal to zero, the trees are no more different than expected by chance; finally if the distance is less than zero, the trees are more different than expected by chance.

\paragraph*{Triplets distance}
The Triplets distance \cite{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees \cite{critchlowthe1996} (see Appendix S2 for calculation details). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between zero when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxa/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa. We normalised this metric in the same way as for the Robinson-Foulds distance resulting in the Normalised Triplets distance.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) using the Bhattacharyya Coefficient \cite{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions \cite{Bhattacharyya} (see Appendix S2 for calculation details). Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients1} for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the "best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.

@@

\subsection*{Individual effects of missing data parameters}
As the amount of missing data increases across all three parameters, our ability to recover the "best" tree topology decreases (Fig.~\ref{Fig_Results-permeth_perparam}).
The Normalised Robinson-Foulds distance is always lower for the Maximum Likelihood trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = 0.69, 0.48 and 0.66 for $M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; Tables S5, S6 and S7). 
However, The Normalised Triplets distance is similar between the Maximum Likelihood trees and the Bayesian consensus trees for all the parameters ($M_{L}$, $M_{F}$ and $M_{C}$) (median Bhattacharrya Coefficient = 0.84, 0.75 and 0.80 for $M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; Tables S5, S6 and S7).

\subsection*{Combined effect of missing data parameters}
As expected, our ability to recover the "best" tree topology is worst when each parameter contains the maximum amount of missing data (i.e. $M_{L}$ = 75\%, $M_{F}$ = 75\% and $M_{C}$ = 75\%), and best when there is no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; Fig.~\ref{Fig_Results-global_perparam}; Tables S2, S3 and S4). Figure ~\ref{Fig_Results-paircomp_within} shows the similarity of distributions of tree metrics in a triangular matrix with the values of each pairwise Bhattacharyya Coefficient coloured according to their values (orange when the distributions overlap completely, Bhattacharyya Coefficient = 1, and blue when they do not, Bhattacharyya Coefficient = 0; Fig.~\ref{Fig_Results-paircomp_within}).

@@

For all combinations of missing data parameters and tree comparison metrics, the Maximum Likelihood bootstrap trees and the Bayesian posterior tree distributions perform very similarly (median Bhattacharrya Coefficient = 0.85 and 0.98, using Normalised Robinson-Foulds distance or Normalised Triplets distance respectively; Table ~\ref{Tab_Results-Difference_methods}). However, these two methods perform worse than the Bayesian consensus trees using Normalised Robinson-Foulds distance (median Bhattacharrya Coefficient = 0 and 0.01, for the Maximum Likelihood bootstrap trees and the Bayesian posterior tree distribution respectively; Table ~\ref{Tab_Results-Difference_methods}; Fig.~\ref{Fig_Results-permeth_perparam} and Fig. S2).

@@

% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
\subsection*{Supplementary informations}
\label{S1_Appendix1}
\noindent{\bf Appendix S1 Tree Inference.} Supplementary information on the Tree Inference methods.

\label{Figure S1}
\noindent{\bf Figure S1 proportion of morphological characters states.}

\label{S2_Appendix2}
\noindent{\bf Appendix S2 Tree Comparisons.} Supplementary information on the Tree Comparison methods.

\label{S2_Appendix3}
\noindent{\bf Appendix S3 Additional results.}.

\label{Figure S2}
\noindent{\bf Figure S2 Effect of increasing missing data on topological recovery using Maximum Likelihood Bootstrap trees (black) and Bayesian posterior tree distribution (grey).}

\label{Figure S3}
\noindent{\bf Figure S3 Effects of missing data on topological recovery using Maximum Likelihood trees.}

\label{Figure S4}
\noindent{\bf Figure S4 Effects of missing data on topological recovery using Maximum Likelihood bootstrap trees.}

\label{Figure S5}
\noindent{\bf Figure S5 Effects of missing data on topological recovery using Bayesian posterior tree distribution.}

\label{Table S1}
\noindent{\bf Table S1 Comparisons between the "best" tree and the "missing data" trees for each different tree inference method.}

\label{Table S2}
\noindent{\bf Table S2 Comparisons between the "best" tree and the "missing data" trees for each different tree inference method for the $M_{L}$ parameter.}

\label{Table S3}
\noindent{\bf Table S3 Comparisons between the "best" tree and the "missing data" trees for each different tree inference method for the $M_{F}$ parameter.}

\label{Table S4}
\noindent{\bf Table S4 Comparisons between the "best" tree and the "missing data" trees for each different tree inference method for the $M_{C}$ parameter.}

\label{Table S5}
\noindent{\bf Table S5 Bhattacharyya Coefficients of the pairwise method comparisons for the $M_{L}$ parameter.}

\label{Table S6}
\noindent{\bf Table S6 Bhattacharyya Coefficients of the pairwise method comparisons for the $M_{F}$ parameter..} Bhattacharyya

\label{Table S7}
\noindent{\bf Table S7 Bhattacharyya Coefficients of the pairwise method comparisons for the $M_{C}$ parameter..} Bhattacharyya

@@

\section*{Author Contributions}
Conceived and designed the experiments: TG NC. Ran the analysis: TG. Analyzed the data: TG. Wrote the paper: TG NC.

>>>>

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

\usepackage{changepage}
\usepackage[utf8]{inputenc}
\usepackage{textcomp,marvosym}
\usepackage{fixltx2e}
\usepackage{amsmath,amssymb}
\usepackage{cite}
\usepackage{nameref,hyperref}
\usepackage[right]{lineno}
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
\usepackage{rotating}
\usepackage{setspace} 
\doublespacing

@@

We inferred the morphological matrix using the R package ape v. 3.0-11 \cite{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (see ~\nameref{S1_Appendix1} and Fig. S1 within). We then ran an independent discrete character simulation for each character using the "true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to another \cite{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \cite{wagner2000,davalosintegrating2014,wrightbayesian2014}.

@@

Each Bayesian tree was estimated using two runs of four chains each for a maximum of 5$\times$$10^7$ generations. We used the following priors for each tree: (i) the "true" tree’s topology as a starting tree (with a starting value for each branch length of one), (ii) an exponential prior on the shape of the gamma distribution of $\alpha$ = 0.5 for both partitions, and (iii) a transition/transversion ratio prior of two sampled from a strong beta distribution ($\beta$(80,40)). We used these priors to speed up the Bayesian estimation process. These priors biased the way the Bayesian process calculated branch lengths by giving non-random starting points and boundaries for parameter estimation however, here we are focusing on the effect of missing data on tree topology and not branch lengths. Even using these priors, it took \texttildelow 140 CPU years to build 50 sets of 125 Bayesian trees (2.30GHz clock speed nodes). The detailed MrBayes parameters are available in the ~\nameref{S1_Appendix1}.

@@

\paragraph*{Robinson-Foulds distance}
The Robinson-Foulds distance \cite{RF1981}, or "path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees \cite{RF1981} (see ~\nameref{S2_Appendix2} for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric by dividing the Robinson-Foulds distance by the mean Robinson-Foulds distance from 1000 random tree comparisons \cite{Bogdanowicz2012}. This allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see ~\nameref{S2_Appendix2} for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the distance is equal to zero, the trees are no more different than expected by chance; finally if the distance is less than zero, the trees are more different than expected by chance.

\paragraph*{Triplets distance}
The Triplets distance \cite{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees \cite{critchlowthe1996} (see ~\nameref{S2_Appendix2} for calculation details). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between zero when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxa/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa. We normalised this metric in the same way as for the Robinson-Foulds distance resulting in the Normalised Triplets distance.

@@

We measured similarities among the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) using the Bhattacharyya Coefficient \cite{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions \cite{Bhattacharyya} (see ~\nameref{S2_Appendix2} for calculation details). Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means. To assess the effect of our missing data parameters, we calculated the Bhattacharyya Coefficient between the distributions of the different metrics scores (Normalised Robinson-Foulds and Normalised Triplets distance) for each pairwise combination of missing data parameters ($M_{L}$, $M_{F}$, $M_{C}$) and parameter states (0\%, 10\%, 25\%, 50\%, 75\%), i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; $M_{L}$ = 10\%, $M_{F}$ = 0\%, $M_{C}$ = 0\% etc. (see Fig.~\ref{Fig_Bhattacharyya_Coefficients1} for more details). This resulted in 7875 pairwise comparisons (a triangular matrix with $3^5$$\times$$3^5$ cells). We performed this procedure separately for each tree inference and uncertainty method. When two combinations of missing data parameters have a similar ability to recover the "best" tree topology the Bhattacharyya Coefficient will be close to one. Conversely, if the two combinations of missing data parameters differ, the Bhattacharyya Coefficient will be close to zero.

@@

\subsection*{Individual effects of missing data parameters}
As the amount of missing data increases across all three parameters, our ability to recover the "best" tree topology decreases (Fig.~\ref{Fig_Results-permeth_perparam}).
The Normalised Robinson-Foulds distance is always lower for the Maximum Likelihood trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = 0.69, 0.48 and 0.66 for $M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; Tables S5, S6 and S7 in ~\nameref{S3_Appendix3}). 
However, The Normalised Triplets distance is similar between the Maximum Likelihood trees and the Bayesian consensus trees for all the parameters ($M_{L}$, $M_{F}$ and $M_{C}$) (median Bhattacharrya Coefficient = 0.84, 0.75 and 0.80 for $M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; Tables S5, S6 and S7 in ~\nameref{S3_Appendix3}).

\subsection*{Combined effect of missing data parameters}
As expected, our ability to recover the "best" tree topology is worst when each parameter contains the maximum amount of missing data (i.e. $M_{L}$ = 75\%, $M_{F}$ = 75\% and $M_{C}$ = 75\%), and best when there is no missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%; Fig.~\ref{Fig_Results-global_perparam}; Tables S2, S3 and S4 in ~\nameref{S3_Appendix3}). Figure ~\ref{Fig_Results-paircomp_within} shows the similarity of distributions of tree metrics in a triangular matrix with the values of each pairwise Bhattacharyya Coefficient coloured according to their values (orange when the distributions overlap completely, Bhattacharyya Coefficient = 1, and blue when they do not, Bhattacharyya Coefficient = 0; Fig.~\ref{Fig_Results-paircomp_within}).

@@

For all combinations of missing data parameters and tree comparison metrics, the Maximum Likelihood bootstrap trees and the Bayesian posterior tree distributions perform very similarly (median Bhattacharrya Coefficient = 0.85 and 0.98, using Normalised Robinson-Foulds distance or Normalised Triplets distance respectively; Table ~\ref{Tab_Results-Difference_methods}). However, these two methods perform worse than the Bayesian consensus trees using Normalised Robinson-Foulds distance (median Bhattacharrya Coefficient = 0 and 0.01, for the Maximum Likelihood bootstrap trees and the Bayesian posterior tree distribution respectively; Table ~\ref{Tab_Results-Difference_methods}; Fig.~\ref{Fig_Results-permeth_perparam} and Fig. S2 in ~\nameref{S3_Appendix3}).

@@

% Include only the SI item label in the subsection heading. Use the \nameref{label} command to cite SI items in the text.
\subsection*{Supporting Information S1: Tree Inference}
\label{S1_Appendix1}
\noindent{\bf Supporting Information S1 Tree Inference.} Supplementary information on the Tree Inference methods. Contains Figure S1.

\subsection*{Supporting Information S2: Tree Comparisons}
\label{S2_Appendix2}
\noindent{\bf Supporting Information S2 Tree Comparisons.} Supplementary information on the Tree Comparison methods.

\subsection*{Supporting Information S3: Additional results}
\label{S3_Appendix3}
\noindent{\bf Supporting Information S3 Additional results.}. Contains Figures S2, S3, S4 and S5 and Tables S1, S2, S3, S4 S5, S6 and S7.

@@

### 2015-04-28 Manuscript/PLoSONEresponse/CoverLetter.tex

Our paper, entitled "Effects of missing data on topological inference using a Total Evidence approach", is to our knowledge, the first to thoroughly analyze the effects of missing data on tree topology in a Total Evidence framework. Using simulations ($>$ 150 CPU years worth), we find that the number of living taxa with morphological data and the overall number of morphological characters, are more important than the amount of missing data in the fossil record for recovering the "best" tree topology. Additionally, we show that Bayesian methods outperform Maximum Likelihood methods, regardless of the amount of missing data.

@@

We feel this paper will be of particular interest to the readers of Systematic Biology as many papers investigating methods for combining living and fossil taxa and the effects of missing data on such analyses (e.g. D\'{a}valos et al. 2014; Pattinson et al. in press, Silvestro et al. 2014, Pyron 2011, Ronquist et al. 2012), have been recently published in the journal.

@@

Suggested reviewers: Liliana D\'{a}valos; David Pattinson; Fredrik Ronquist; Nicolas Salamin; Tanja Stadler.

>>>>

Our research article, entitled "Effects of missing data on topological inference using a Total Evidence approach", is to our knowledge, the first to thoroughly analyze the effects of missing data on tree topology in a Total Evidence framework. Using simulations ($>$ 150 CPU years worth), we find that the number of living taxa with morphological data and the overall number of morphological characters, are more important than the amount of missing data in the fossil record for recovering the "best" tree topology. Additionally, we show that Bayesian methods outperform Maximum Likelihood methods, regardless of the amount of missing data.

@@

We would like to submit this as a research article. We have had no prior interactions with PLoS regarding this article.

As academic editors we suggest Matthew C. Mihlbachler, Patrick O'Grady and Nicholas Pyenson. We have no opposed reviewers or editors.

@@

### 2015-04-28 Manuscript/PLoSONEresponse/TEM_PLoS_response.tex

\paragraph*{Robinson-Foulds distance}
The Robinson-Foulds distance \cite{RF1981}, or ``path difference", measures the number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees \cite{RF1981} (see ~\nameref{S2_Appendix2} for calculation details). This metric is bounded between one, when the two trees are identical, and $n-2$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric by dividing the Robinson-Foulds distance by the mean Robinson-Foulds distance from 1000 random tree comparisons \cite{Bogdanowicz2012}. This allows us to compare the metric for any number of taxa by scaling it with the expected value of the metric when comparing two random trees (see ~\nameref{S2_Appendix2} for the calculation details). Therefore, when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the distance is equal to zero, the trees are no more different than expected by chance; finally if the distance is less than zero, the trees are more different than expected by chance.

\paragraph*{Triplets distance}
The Triplets distance \cite{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees \cite{critchlowthe1996} (see ~\nameref{S2_Appendix2} for calculation details). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between zero when the two trees are identical and $\binom{n}{4}$ (for two trees with $n$ taxa) when there is no shared taxa/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa. We normalised this metric in the same way as for the Robinson-Foulds distance resulting in the Normalised Triplets distance.

@@

\begin{figure}[h]
\caption{{\bf Bhattacharyya Coefficient calculation outline 1.}
 A, B and C are distributions of tree similarity metrics (Normalised Robinson-Foulds or Normalised Triplets distances) for any combination of missing data parameters (e.g. $M_{L}$ = 10\%, $M_{F}$ = 50\%, $M_{C}$ = 75\%). The Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity metrics between two combinations of missing data parameters, for example, BC(A,B) is the probability of overlap between the distributions A and B. Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means.}
\label{Fig_Bhattacharyya_Coefficients1}
\end{figure}

@@

\begin{figure}[h]
\caption{{\bf Bhattacharyya Coefficient calculation outline 2.}
 A and B are distributions of tree similarity metrics (Normalised Robinson-Foulds or Normalised Triplets distances) for any combination of missing data parameters (e.g. $M_{L}$ = 10\%, $M_{F}$ = 50\%, $M_{C}$ = 75\%). \textbf{(x)} and \textbf{(y)} are two different tree inference methods (e.g. Maximum Likelihood or Bayesian). The Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity metrics between two methods for the same combination of missing data parameters, for example, BC($A_{x}$,$A_{y}$) is the probability of overlap of the distribution A for methods $x$ and $y$. Note that this is equivalent to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means.}
\label{Fig_Bhattacharyya_Coefficients2} 
\end{figure}

>>>>

\paragraph*{Robinson-Foulds distance}
The Robinson-Foulds distance \cite{RF1981}, or ``path difference", measures the difference between the number of clades and the twice number of shared clades across two trees. The metric reflects the distance between the distributions of tips among clades in the two trees \cite{RF1981} (see ~\nameref{S2_Appendix2} for calculation details). This metric is bounded between zero, when the two trees are identical, and $2(n-2)$ (for two trees with $n$ taxa) when there is no shared clade in the two trees. This metric is sensitive to minor changes in clade conservation: if the trees are composed of two clades of three taxa (\textit{(((a,b),c),((d,e),f))}), the swapping of any two taxa will lead to a maximal score of the Robinson-Foulds distance indicating poor tree similarity. We normalised this metric following Bogdanowicz \textit{et al.} \cite{Bogdanowicz2012}'s Normalised Tree Similarity (NTS) method. This methods scales %not sure if scale here?
any tree comparison metric using the mean distance between 1000 random trees (see ~\nameref{S2_Appendix2} for the calculation details). This method is a generalisation of the topological accuracy method \cite{Price2010} allowing to compare topological differences between any tree with any tree comparison metric. In practice when the Normalised Robinson-Foulds distance between two trees is equal to one, the trees are identical; if the distance is equal to zero, the trees are no more different than expected by chance; finally if the distance is less than zero, the trees are more different than expected by chance.

\paragraph*{Triplets distance}
The Triplets distance \cite{dobson1975triplets} measures the number of sub-trees made up of three taxa that differ between two trees \cite{critchlowthe1996} (see ~\nameref{S2_Appendix2} for calculation details). This metric measures the position of each taxon and clade in relation to its closest neighbours. It is bounded between zero when the two trees are identical and $\binom{n}{3}$ (for two trees with $n$ taxa) when there is no shared taxa/clade position in the two trees. Therefore this metric sensitive to the conservation of wildcard taxa. We normalised this metric in the same way as for the Robinson-Foulds distance resulting in the Normalised Triplets distance.

@@

\begin{figure}[h]
\caption{{\bf Bhattacharyya Coefficient calculation outline 1.}
 A, B and C are distributions of tree similarity metrics (Normalised Robinson-Foulds or Normalised Triplets distances) for any combination of missing data parameters (e.g. $M_{L}$ = 10\%, $M_{F}$ = 50\%, $M_{C}$ = 75\%). The Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity metrics between two combinations of missing data parameters, for example, BC(A,B) is the probability of overlap between the distributions A and B. Note that this is similar to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means.}
\label{Fig_Bhattacharyya_Coefficients1}
\end{figure}

@@

\begin{figure}[h]
\caption{{\bf Bhattacharyya Coefficient calculation outline 2.}
 A and B are distributions of tree similarity metrics (Normalised Robinson-Foulds or Normalised Triplets distances) for any combination of missing data parameters (e.g. $M_{L}$ = 10\%, $M_{F}$ = 50\%, $M_{C}$ = 75\%). \textbf{(x)} and \textbf{(y)} are two different tree inference methods (e.g. Maximum Likelihood or Bayesian). The Bhattacharyya Coefficient (BC) is the overlap of the distribution of tree similarity metrics between two methods for the same combination of missing data parameters, for example, BC($A_{x}$,$A_{y}$) is the probability of overlap of the distribution A for methods $x$ and $y$. Note that this is similar to performing a two-sided t-test, but we use the Bhattacharyya Coefficient here because we are comparing distributions not means.}
\label{Fig_Bhattacharyya_Coefficients2} 
\end{figure}
